---
title   : 'Data Science for Linguists'
subtitle: 'The linear model:</br>.lightgrey[Multiple regression]'
author  : "Joseph V. Casillas, PhD"
date    : "Rutgers University</br>Spring 2021</br>Last update: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: assets
    css: ["hygge", "rutgers", "rutgers-fonts"]
    nature:
      beforeInit: ["https://www.jvcasillas.com/ru_xaringan/js/ru_xaringan.js", "https://platform.twitter.com/widgets.js"]
      highlightStyle: default
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
    includes:
      in_header: "../../assets/partials/header.html"
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra-all-the-things, echo=FALSE}
xaringanExtra::use_xaringan_extra(
  c("tile_view", "panelset", "editable", "animate", "tachyons", "webcam")
)
```

```{r, 'helpers', echo=FALSE, message=F, warning=F}
source(here::here("assets", "scripts", "helpers.R"))
source(here::here("05_lm", "03_mrc", "assets", "scripts", "mrc.R"))
```

```{r, load_refs, echo=FALSE, cache=FALSE, warning=F, message=F}
bib <- ReadBib(here("assets", "bib", "ds4ling_refs.bib"), check = FALSE)
ui <- "- "
```



```{r, child="part1"}
```

```{r, child="part2", eval=F}
```

---
exclude: true

`r AutoCite(bib, c("wickham2016r", "qass22_ch3", "qass57_ch2", "qass57_ch5", "qass50_ch2", "qass50_ch4"))`. 

---
layout: false
class: title-slide-final, left

# References

```{r, results='asis', echo=FALSE, eval=TRUE, cache=FALSE}
PrintBibliography(bib, end = 5)
```

---
layout: false
class: title-slide-final, left

# References (cont.)

```{r, results='asis', echo=FALSE, eval=TRUE, cache=FALSE}
PrintBibliography(bib, start = 6)
```

Figueredo, A. J. (2013). Multiple Regression. *Statistical 
Methods in Psychological Research*.

Figueredo, A. J. (2013). Strong Inference. *Statistical 
Methods in Psychological Research*.

Figueredo, A. J. (2013). Empirical Selection of Variables. *Statistical 
Methods in Psychological Research*.
