---
title   : 'Data Science for Linguists'
subtitle : 'The linear model: [Assumptions, diagnotics, and interpretation]{style="color: #888; font-size: 0.9em;"}'
author   : "Joseph V. Casillas, PhD"
institute: "Rutgers University<mybr>Spring 2025<br>Last update: `r Sys.Date()`"
---

```{r}
#| label: load-helpers
#| echo: false 
#| message: false 
#| warning: false

source(here::here("assets", "scripts", "helpers.R"))
read_chunk(here("05_lm", "02_assumptions_diagnostics", "index_files", "scripts", "assumptions_diagnostics.R"))
```

# {.transition visibility="uncounted"}

{{< tweet user=ChelseaParlett id=1352717277976694784 >}}
<!-- regression the movie tweet --> 


# Assumptions {.transition}

---

## Assumptions {background-image="./index_files/img/assumptions.png" background-size="350px" background-position="95% 50%" data-menu-title="General assumptions"}

### We make a lot of assumptions

::: {.columns}
::: {.column style="font-size: 0.7em; width: 75%;"}
<ru-blockquote>
[1.]{color="lightblue"} the act of assuming or taking for granted  
[2.]{color="lightblue"} a hypothesis that is taken for granted  
[3.]{color="lightblue"} a thing that is accepted as true or as certain to happen, without proof
</ru-blockquote>
:::
:::

::: {.columns}
::: {.column .closelist .absolute bottom="10%" style="font-size: 0.7em;"}
- About the world
- About others
- About reactions
- About academia
:::

::: {.column .closelist .absolute bottom="10%" right="0" style="font-size: 0.7em;"}
- The sun will rise tomorrow
- People are generally X
- They will find my joke funny
- Reviewer 2 is evil
:::
:::

---

## Assumptions {background-image="https://raw.githubusercontent.com/jvcasillas/media/master/rstats/memes/lm_assumptions.png" background-size="350px" background-position="95% 50%" data-menu-title="Statistical assumptions"}

### Statistical assumptions

::: {.columns}
::: {.column style="font-size: 0.9em; width: 75%;"}
- Every model has assumptions 
- They need to be met for the model to work properly and be trustworthy
- It is not standard practice to report whether all assumptions have been met when writing up results... [but it should be]{.emph}
- We will make a habit of assessing models and incorporating the relevant information in prose
:::
:::

---

## Assumptions {background-image="https://raw.githubusercontent.com/jvcasillas/media/master/rstats/memes/lm_assumptions.png" background-size="350px" background-position="95% 50%" data-menu-title="Regression assumptions"}

### Regression assumptions

We can break the assumptions up into 3 areas: 

1. Model specification
2. Measurement error
3. The error term



# Model specification {.transition}

---

## Model specification {data-menu-title="Specification error"}

### Specification error

There should be no specification error

- The relationship between x<sub>i</sub> and y<sub>i</sub> is linear: 
$$y_i = \alpha + \beta x_i + \epsilon _i$$
- Including irrelevant variables 🤷🏽
- Omitting relevant variables 👎🏽

::: {.fragment}
### Importance: [**HIGH**]{.emph}
:::

---

## Model specification {data-menu-title="Including an irrelevant variable" background-image="./index_files/img/sin.png" background-size="400px" background-position="95% 50%"}

### Including an irrelevant variable (sin of commission): 🤷🏽

::: {.columns}
::: {.column style="font-size: 0.85em; width: 70%;"}
- Similar to a Type I error
- Adds to the error of prediction, but it does not bias the parameter estimates 
- If model includes multiple predictors, other estimates are not influenced by the irrelevant variable
- Though the parameter estimates won't be biased, the standard error terms 
around each beta weight will increase (this affects t-ratio and p-value)
:::
:::

---

## {background-image="./index_files/img/sin.png" background-size="400px" background-position="95% 50%"}

[Model specification]{.emph .p-font style="font-size: 1.85em;"}

[Including an irrelevant variable (sin of commission): 🤷🏽]{.p-font style="font-size: 1.3em; color: #666666"}

::: {.columns}
::: {.column style="font-size: 0.85em; width: 70%;"}
- For example, imagine a situation in which you model employee performance as a function of a bunch of good predictors
- Adding the variable of astrological horoscope sign will not bias the other estimates
- You are considering something that doesn't matter, i.e., astrology
:::
:::

---

## Model specification {background-image="./index_files/img/sin.png" background-size="400px" background-position="95% 50%" data-menu-title="Excluding a relevant variable"}

### Excluding a relevant variable (sin of omission): 👎🏽

::: {.columns}
::: {.column style="font-size: 0.85em; width: 70%;"}
- Similar to a Type II error
- This is much more serious!
- This will bias your parameter estimates
- You are leaving something important out (variance left unexplained)
:::
:::

---

## {background-image="./index_files/img/sin.png" background-size="400px" background-position="95% 50%"}

[Model specification]{.emph .p-font style="font-size: 1.85em;"}

[Excluding a relevant variable (sin of omission): 👎🏽]{.p-font style="font-size: 1.3em; color: #666666"}

::: {.columns}
::: {.column style="font-size: 0.85em; width: 70%;"}
- For example, if you want to model RT as a function of working memory but you forget to look at age
- Then your prediction might be wrong
- You don’t know which way you are biasing things either!
- You could be either overestimating or underestimating the other regression parameters
:::
:::

---

## {background-image="./index_files/img/sin.png" background-size="400px" background-position="95% 50%" data-menu-title="What to do?"}

[Model specification]{.emph .p-font style="font-size: 1.85em;"}

[What to do]{.p-font style="font-size: 1.3em; color: #666666"}

#### [After including an irrelevant predictor:]{color="black"}

::: {.closelist style="font-size: 0.8em;"}
- Drop the irrelevant variable
- Not much of a problem
:::

<mybr>

::: {.fragment}
#### [After excluding a relevant predictor:]{color="black"}

::: {.closelist style="font-size: 0.8em;"}
- Look at data again to see if it is in there
- Hope and Pray!
- If it’s not in you data, you are SOL
- You need to repeat the study!
- There is a problem in your design and there is no mathematical solution to fix this problem
:::
:::



# Measurement error {background-image="./index_files/img/measure0.png" background-size="contain" background-position="100% 50%"}

---

## Measurement error {data-menu-title="Measuring is hard" background-image="https://media.tenor.com/AFJK9BNwiZIAAAAM/harry-styles-tape-measure.gif" background-size="300px" background-position="98% 10%"}

### Measuring things is hard

::: {.columns}
::: {.column .fragment .absolute left="0" top="35%" width="55%"}
- How far away is that lion?
- How much milk do I need?
- Can I park there?
- How much weight did I gain?
:::

::: {.column .absolute right="0" top="30%" width="10%"}
:::

::: {.column .fragment .absolute right="0" top="35%" width="35%"}
- Human error
- Specificity
- Operationalizations
- Repeatability
:::
:::

---

## Measurement error {data-menu-title="Minimize measurement error" background-image="./index_files/img/measure1.png" background-size="contain" background-position="110% 50%"}

### There should be no measurement error

::: {.columns}
::: {.column .absolute top="35%"}
- Variables x<sub>i</sub> and y<sub>i</sub> should be as accurate as possible
:::
:::

::: {.absolute bottom="5%"}
[Importance: [high]{.emph}]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}
:::

---

## Measurement error {data-menu-title="Consequences" background-image="https://media4.giphy.com/media/v1.Y2lkPTc5MGI3NjExNmV2M2xiZXBhemV3MGdhNzFreWpuNXg1Mmd5YnB4MDh0bnRwNWhnZyZlcD12MV9pbnRlcm5hbF9naWZfYnlfaWQmY3Q9Zw/k5lj4s1qxaSyI/giphy.webp" background-size="500px" background-position="98% 50%"}

### Consequences of measurement error

::: {.columns}
::: {.column .incremental style="width: 65%; font-size: 0.9em;"}
- Your models will estimate the parameters you specify
- But you won't know if the estimates are reliable
- Large degree of unquantifiable uncertainty
- Cannot "safely" do NHST
- Your research may be reproducible, but probably will not be replicable
:::
:::

<!-- anchorman gif, 60% of the time it works every time -->

---

## Measurement error {data-menu-title="What to do?" background-image="./index_files/img/measure2.gif" background-size="500px" background-position="98% 50%"}

### What you can do

::: {.columns}
::: {.column .incremental style="width: 65%; font-size: 0.9em;"}
- The solution to this problems revolves around proper planning a priori
- Operationally define every feasible aspect of *how* and *what* you will measure
  - Use previous literature, if available
  - Preregister decisions
- Repeated measures, if possible
- Random, 3rd party quality checks
- Measurement error models, if possible
:::
:::

<!-- anchorman gif, 60% of the time it works every time -->

---

# The error term {.transition}

---

## The error term {data-menu-title="About the error term"}

### About the error term

The error term should meet the following characteristics...

- Mean = 0
- Homoskedasticity
- No autocorrelation
- Predictor not correlated with error term
- Error is normally distributed 

---

## About the error term {data-menu-title="Mean of 0"}

### Mean of 0

If the mean of the error term deviates far from 0...

- Intercept can be biased
- Slope estimates will not be biased

[Importance: [low]{color="blue"}... unless you are interested in the intercept.]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}

---

## About the error term {data-menu-title="Homoskedasticity"}

### Homoskedasticity

::: {.closelist}
- Variance around predicted values should be consistent
- Common simple inspection is to look at scatter plot of fitted vs. x-values (should look like blob with no interesting patterns)
- If variance is heteroskedastic you will typically see fan-like patterns
- Will not bias parameter estimates
- Will increase confidence intervals, thus affects t-ratios and p-values
:::

[Importance: [medium/high]{color="orange"}]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}

---

## {background-color="black" visibility="uncounted" .center}

<center>
<iframe height="390" width= "520" src="https://www.tiktok.com/player/v1/7076276399505558827?music_info=1&description=1" allow="fullscreen" title="test"></iframe>
</center>

::: footer
<https://www.tiktok.com/@chelseaparlettpelleriti/video/7076276399505558827>
:::

<!-- homoskedasticity -->

---

## About the error term {data-menu-title="Autocorrelation"}

### No autocorrelation

::: {style="font-size: 0.8em;"}
- If the residuals are autocorrelated, it means that the prediction error of a given observation depends on that of the previous observation
- This shows up as a clear unexplained pattern in the y variable 
- Most common in repeated measures and longitudinal data
- Will not bias parameter estimates
- Will affect confidence intervals, t-ratios, p-values
- Increased chance of Type II error
:::

[Importance: [High]{.emph}, but uncommon in standard regression]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}

---

## About the error term {data-menu-title="Predictor correlation with error term"}

### Predictor(s) should not be correlated with error term

- Typically the result of omitting a relevant variable (sin of omission)
- Will bias parameter estimates
- Solution: include missing variable

[Importance: [high]{.emph}]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}

---

## About the error term {data-menu-title="Normally distributed errors"}

### Error (residuals) should be normally distributed

::: {.incremental .closelist style="font-size: 0.8em;"}
- There is no a priori reason for error to be anything but normally distributed 
- It should become standard practice for you to examine your residuals to see if they are normally distributed or not
- If they aren't normally distributed there is a substantial possibility of Type II error
- If this is the case, then the residuals may not be "pure" error, and you may have omitted a relevant variable from the equation that is making the residuals not be normally distributed
- In other words, the residuals may contain systematic variance that can still be explained by something else
- You cannot conclude with 100% certainty that a Type II error has been 
committed, but you might strongly suspect it
:::

[Importance: [high]{.emph}]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}







# Summing up {.smaller}

::: {.closelist}
- Model specification
  - The relationship between x<sub>i</sub> and y<sub>i</sub> is linear
  - Including irrelevant variables 
  - Omitting relevant variables
:::

- Measurement error

::: {.closelist}
- The error term
  - Mean = 0
  - Homoskedasticity
  - No autocorrelation
  - Predictor not correlated with error term
  - Error is normally distributed
:::




# Diagnostics Examples {.transition}

---

## Diagnostics {.smaller data-menu-title="More about residuals"}

### Remember those "residuals"?

::: {.closelist}
- A residual represents prediction error in our model with regard to a single point. 
  - Our `mtcars` model predicted that a 6 ton car should get 5.22 mpg. 
  - If in reality it gets 10.22 mpg, then the residual for that specific data point would be 5 mpg. 
:::

::: {.fragment}
- Remember: All models are wrong. 
:::

::: {.closelist .fragment}
- Our models will always be off by at least a little bit for most of the observations we try to fit. 
- Nonetheless, it is good practice to examine the residuals of your models in order to help assess how well it fits your data.
- Specifically, we need to make sure that there aren't any unexpected patterns because that would suggest that our model does not properly fit our data.
:::

---

## {background-image="./index_files/img/sick-computer.png" background-size="400px" background-position="95% 50%"}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

[Considerations for model diagnostics]{.p-font style="font-size: 1.2em; color: #666666"}

<br>

[1. Model assumptions]{.p-font style="font-size: 1.2em;"}

[2. Outliers]{.p-font style="font-size: 1.2em;"}

---

## {data-menu-title="Assumption of linear relationship" .smaller}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### The relationship between x<sub>i</sub> and y<sub>i</sub> is linear.

1. Double check linear specification
2. Eyeball it

```{r}
#| label: fit_mods
data(assumptions_data)
mod1 <- lm(y ~ x, data = assumptions_data)
mod2 <- lm(y_quad ~ x, data = assumptions_data)
```

<mybr>

::: {.columns}
::: {.column}
```{r}
#| label: assumptions_linear_plot
#| fig-asp: 0.8
assumptions_data |> 
  ggplot() + 
  aes(x = x, y = y) + 
  geom_point(
    pch = 21, color = 'darkgrey', fill = 'darkred', 
    stroke = 1, size = 4
  ) + 
  ds4ling_bw_theme(base_size = 24)
```
:::

::: {.column}
```{r}
#| label: assumptions_quadratic_plot
#| fig-asp: 0.8
assumptions_data |> 
  ggplot() + 
  aes(x = x, y = y_quad) + 
  geom_point(
    pch = 21, color = 'darkgrey', fill = 'darkred', 
    stroke = 1, size = 4
  ) + 
  ds4ling_bw_theme(base_size = 24)
```
:::
:::

---

## {data-menu-title="Mean of residuals = 0" .smaller}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### The mean of residuals is zero

1. Fit model
2. Get residuals
3. Test manually

<mybr>

::: {.columns}
::: {.column}
```{r}
#| label: model-ex-code1
#| echo: true
# Linear relationship
# Fit model and print summary
mod1 <- lm(y ~ x, data = assumptions_data)
summary(mod1$residuals)
```
:::

::: {.column}
```{r}
#| label: model-ex-code2
#| echo: true
# Non-linear relationship
# Fit model and print summary# Fit mod
mod2 <- lm(y_quad ~ x, data = assumptions_data)
summary(mod2$residuals)
```
:::
:::

---

## {data-menu-title="Homoskedasticity of residuals"}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### Homoskedasticity of residuals

```{r}
#| label: homoskedasticity_linear
#| fig-asp: 0.45
autoplot(mod1, which = c(1, 3)) + 
  ds4ling_bw_theme(base_size = 18)
```

::: aside
mod1
:::

---

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

[Homoskedasticity of residuals]{.p-font style="font-size: 1.2em; color: #666666"}

```{r}
#| label: homoskedasticity_quadratic
#| fig-asp: 0.45
autoplot(mod2, which = c(1, 3)) + 
  ds4ling_bw_theme(base_size = 18)
```

::: aside
mod2
:::

---

## {data-menu-title="No autocorrelation of residuals"}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### No autocorrelation of residuals

::: {.columns}
::: {.column}
1. Visual inspection
2. Durbin-Watson test

```{r}
#| label: linear_durbin_watson
#| comment: ''
# formal test: Durbin-Watson test
lmtest::dwtest(mod1)
```
:::

::: {.column}
```{r}
#| label: linear_autocorrelation
#| fig-asp: 0.8
#| out-extra: "style=float:right"
# visual inspection
acf(mod1$residuals) 
```
:::
:::

::: aside
mod1
:::

---

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

[No autocorrelation of residuals]{.p-font style="font-size: 1.2em; color: #666666;"}

::: {.columns}
::: {.column}
1. Visual inspection
2. Durbin-Watson test

```{r}
#| label: quadratic_durbin_watson
#| comment: ''
# formal test: Durbin-Watson test
lmtest::dwtest(mod2)
```
:::

::: {.column}
```{r}
#| label: quadratic_autocorrelation
#| fig-asp: 0.8
# visual inspection
acf(mod2$residuals) 
```
:::
:::

::: aside
mod2
:::

---

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

[No autocorrelation of residuals - Correction]{.p-font style="font-size: 1.2em; color: #666666;"}

::: {.columns}
::: {.column .small-code}
```{r}
#| label: auto_cor_fix
#| fig-asp: 0.8
#| echo: true
autocor_data <- data.frame(
  assumptions_data, 
  resid_mod2 = mod2$residuals
)

autocor_data1 <- DataCombine::slide(
  autocor_data, 
  Var = "resid_mod2", 
  NewVar = "lag1", 
  slideBy = -1
)
autocor_data2 <- na.omit(autocor_data1)
mod2_fix <- lm(y_quad ~ x + lag1, data = autocor_data2)
```

```{r}
#| label: autocor-durbin-watson
#| comment: ''
lmtest::dwtest(mod2_fix)  # formal test: Durbin-Watson test
```
:::

::: {.column}
```{r}
#| label: auto-cor-fix-plot
#| fig-asp: 0.8
acf(mod2_fix$residuals)
```
:::
:::

::: footer
mod2
:::

---

## {data-menu-title="Predictors and residuals are uncorrelated"}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### Predictors and residuals are uncorrelated

::: {.closelist .small-code}
- Test for correlation
- Think about your study

```{r}
#| label: pred-resid-corr-test-linear
# do correlation test 
cor.test(assumptions_data$x, mod1$residuals)
```
:::

::: aside
mod1
:::

---

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

[Predictors and residuals are uncorrelated]{.p-font style="font-size: 1.2em; color: #666666;"}

::: {.closelist .small-code}
- Test for correlation
- Think about your study

```{r}
#| label: pred_resid_corr_test_quadratic
# do correlation test 
cor.test(assumptions_data$x, mod2$residuals)
```
:::

::: aside
mod2
:::

---

## {data-menu-title="Normality of residuals" .smaller}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### Normality of residuals

- QQplots

```{r}
#| label: normal_resids_linear-quadratic
#| fig-asp: 0.45
#| fig-align: 'center'
ap1 <- autoplot(mod1, which = 2) + 
  ds4ling_bw_theme(base_size = 18)

ap2 <- autoplot(mod2, which = 2) + 
  ds4ling_bw_theme(base_size = 18)

ap1 + ap2
```





# Dealing with influential data points {data-menu-title="Influential data points" .center}

---

## sdkjfaj


